---
project:
  type: book
  output-dir: .
book:
  title: "Thesis Outline"
  subtitle: "Thesis Title: Advancing Snow Accumulation Models in Needleleaf Forests"
  author: "Alex Cebulski"
  date: today
  date-format: long
  chapters:
    - chapters/02-review-paper.qmd
bibliography: ../../papers/bibtex/better-bibtex.bib
csl: '../../papers/bibtex/csl-styles/apa-disable-disambiguate-add-names.csl'
---

---

*Keywords:* Snow, Hydrology, Climate, Forest, Interception, Ablation

*Supervisor:* Dr. John Pomeroy

*Degree:* Doctor of Philosophy

*Academic Unit:* Department of Geography and Planning

---

# Abstract {.unnumbered}

Interception and ablation of snow in needleleaf canopies regulates the quantity, timing, and phase of precipitation that reaches the ground in cold region forests. Sparse observations have limited understanding of these processes and hindered their representation in hydrological and land-surface models. Consequently, simulations of subcanopy snow accumulation have shown variable accuracy across different climates and canopy structures, and diagnosing the causative processes that partition snowfall in needleleaf forests remains difficult. This study aims to better understand snow interception and canopy snow ablation processes, which act collectively to govern subcanopy snow accumulation in forested environments. The specific objectives of this thesis are: 1) Evaluate the suitability of existing snow interception and ablation parameterisations for application in needleleaf forests with differing canopy structure and meteorology. 2) Determine how new snow interception and ablation parameterisations could enhance the representation of processes important for subcanopy snow accumulation. To achieve these objectives, observations of forest-snow processes and subcanopy snow accumulation were collected to evaluate the theories underlying existing parameterisations, derive new relationships, and validate a newly proposed model.

New observations of initial snow interception, collected when ablation processes were minimal, showed that canopy density is the primary factor influencing subcanopy snow throughfall. Contrary to existing theories, no relationships were found with canopy snow load or air temperature. Canopy density was best represented by a snow-leaf contact area metric, which accounts for increasing contact area with more horizontal snowfall trajectories. This metric was highly sensitivity to wind speed in sparse canopies, increasing by up to an order of magnitude with a 1 m s^-1^ wind speed. A new parameterisation was developed that calculates throughfall as a function of snowfall and snow-leaf contact area based on these observations. This approach is consistent with some rainfall interception studies, which also separate canopy loading and ablation processes, and calculate interception as a function of canopy cover. 

Observations showed that canopy snow unloading was strongly associated with snow load, wind shear stress, and canopy snowmelt, but not with air temperature or sublimation. A new canopy snow ablation model was developed based on these associations and their impact on the canopy snow energy and mass balance. The model improved simulation of canopy snow load relative to previous approaches, especially for melt- and wind-dominated ablation events. The improved performance in representing canopy snow load compared to existing models results from including energy balance-based melt and dry snow unloading relationships with snow load and wind shear stress. The inclusion of both melt and dry snow unloading processes in the new model also led to more accurate partitioning of snowfall to the atmosphere versus the ground compared to existing approaches across a wide range of meteorologies.

Validation of the revised canopy snow mass and energy balance was conducted at four needleleaf forests characterised by differing tree species, canopy structure, and meteorological conditions. The new physically based approach reduced error in simulating subcanopy snow water equivalent (SWE) compared to an existing model. Improved process representation also enabled clearer process diagnosis on the influence of vegetation on snow accumulation. At two cold, low-snowfall sites, roughly half of annual snowfall was lost to the atmosphere via sublimation of intercepted snow. A cold wind-exposed site with higher annual snowfall had increased unloading rates and reduced atmospheric losses. In contrast, at a temperate-maritime site, nearly half of annual snowfall melted within the canopy, and combined with melt-induced unloading led to the lowest fraction of snowfall lost through canopy snow sublimation.

This research identified key limitations in existing models using in-situ process-level observations of snow interception and ablation. New parameterisations of the canopy snow mass and energy balance are introduced to more explicitly represent the processes that govern snowfall partitioning in needleleaf canopies. This new approach led to more accurate simulations of canopy snow load and subcanopy SWE across a broad range of meteorologies and forest types and also improved process diagnosis, though further validation is required across a diverse range of sites. 

# Introduction

Melt of the seasonal mountain snowpack is an important source of streamflow across the globe, crucial for downstream ecosystems, energy production, and agricultural irrigation [@Viviroli2007; @Derksen2019]. The high albedo and energy required to melt snow also provides a large-scale cooling effect for our planet by reflecting most of the incoming solar radiation and dissipating energy during spring melt [@Henderson2018]. Despite this importance, there is uncertainty in estimates of snow accumulation, especially in mountain forests, due redistribution of snow by wind and forest canopy [@Rutter2009; @Krinner2018; @Ellis2010]. In western Canada, diverse topography, climate and canopy structure results in highly variable snow accumulation, which presents a challenge in predicting the quantity of snow available for streamflow and downstream water resources [@Pomeroy1995; @Ellis2010]. In the Northern Hemisphere, over 50% of snowmelt-dominated basins are covered by needleleaf forest [@Kim2017]. This extensive canopy coverage reduces the amount of snow that is available for streamflow through interception of snowfall and subsequent ablative processes [@Hedstrom1998; @Essery2003; @Ellis2010]. Intercepted snow in the canopy is subjected to higher rates of sublimation compared to subcanopy snow due to greater surface area, turbulent energy exchange, and solar exposure [@Pomeroy1998b]. Across the Northern Hemisphere, researchers estimate that 25 to 45% of annual snowfall may be lost to the sublimation of intercepted snow from the canopy [@Essery2003]. However, the time that snow resides in the canopy and is subject to sublimation is dependent on rates of unloading, melt, drip, and resuspension of snow. These processes lead to the spatial and temporal diversity in subcanopy snowpack accumulation and contribute to the higher uncertainty for the forested domain of hydrological models [@Rutter2009; @Krinner2018].

The high variability of snow accumulation in forests is not well represented due to a sparse and unrepresentative network of in situ observations, which are mostly located in clearings [@Vionnet2021]. Moreover, observing snow accumulation under forest canopies at large extents remains uncertain with current remote sensing technologies [@Rittger2020]. There is a need for reliable models of snow redistribution by forest and wind to estimate snow accumulation in mountains [@Pomeroy2007; @Rutter2009; @Clark2015]. Existing snow interception parameterisations have been developed for both warm maritime [@Storck2002; @Andreadis2009] and cold continental [@Hedstrom1998; @Ellis2010; @Roesch2001] climates characterized by dense forest canopy. Accurate simulations of forest snow accumulation is often achieved if the parameterisations are applied in similar climates to where they were developed [@Lundquist2021; @Roth2019; @Rasouli2019b] or if they are combined into a hybrid parameterisation and assessed at global and regional scales in a wide range of climates [@Essery2003; @Gelfan2004]. Although accurate performance has been achieved across different climates in some studies [@Essery2004; @Gelfan2004], other snow model comparisons [@Krinner2018; @Rutter2009] have shown reduced performance. The decision in earth system models to use parameterisations derived in warm or cold climates is often based on a simple temperature-based step function [@Essery2003; @Gelfan2004] and may require modification to better represent more transitional climates and forest types. The omission or simplified representation of processes and reliance on empirical calibrations likely contribute to model uncertainty when applied in climates and forests where other processes become important [@Rutter2009; @Krinner2018; @Lumbrazo2022; @Roth2019; @Moeser2015; @Lundquist2021]. 
  
Some of the processes currently not encoded in snow interception models include wind redistribution of snow during interception, rain on intercepted snow, hoarfrost, rime ice and the cohesion and adhesion of snow in the canopy. @Roesch2001 and @Bartlett2015 created a wind unloading function for canopy snow that is based on measurements of snow albedo from @Betts1997 as an index for canopy snow ablation (i.e., unloading, melt, and sublimation) in a low wind environment of the boreal forest in Saskatchewan. However, @Pomeroy1996 found the relationship between albedo and intercepted snow load was not discernible for the same forest as the @Betts1997 study. Moreover, it is unclear the influence of sublimation and melt processes on the canopy snow ablation proxy measurement. @Pomeroy1996 used careful radiation measurements and snow loads from a weighed suspended tree and challenge the @Betts1997 observations the @Roesch2001 parameterisation is based on. @Lumbrazo2022 provide a more recent assessment of canopy snow wind redistribution and provide the the first uncalibrated test of the @Roesch2001 wind induced unloading parameterisation. The results from @Lumbrazo2022 generally showed poor performance in mountain forests of Colorado and Washington. @Lumbrazo2022 additional process investigation on the influence of wind on canopy snow interception is required to better simulate mountain forest SWE accumulation. 

The calculation of canopy snow melt is often oversimplified in many snow interception models. For example, [@Pomeroy2007; @Ellis2010; @Essery2016] use empirical time-based melt calculations to melt snow intercepted in the canopy. While @Clark2020 and @Verseghy2017 use a physically based energy balance model to melt snow in the canopy, both models assume that the surface temperature of vegetative elements are in equilibrium with snow intercepted in the canopy. This simplification likely leads to a positive bias in canopy snow melt predictions as the vegetative elements are known to have higher temperatures than snow clumps [@Pomeroy2009]. Once the canopy snow is at 0 °C and is melting, intercepted snow will reach the ground as clumps of wet snow or will drip as liquid water to the ground [@Storck2002]. The distinction between wet snow or liquid meltwater has important implications for the energy balance of the snowpack beneath the canopy, as liquid meltwater can add a large amount of advective energy. However, predicting the ratio of solid to liquid water unloading due to canopy snow melt remains an area of uncertainty [@Roesch2001; @Storck2002].

Future climate change is expected to change the dominant hydrological processes in mountain forests [@He2021; @Dettinger2014; @Viviroli2011; @Fang2020b]. A report by Environment and Climate Change Canada [@Bush2019], suggest an increase in air temperature of 2 °C from the 1986-2005 reference period by 2050 for the low emission scenario and 6 °C for the high emission scenario by the late 21^st^ century. @Bush2019 project an increase in annual mean precipitation over all of Canada during the second half of the 21^st^ century, with a larger increase in northern Canada. As the climate warm canopy snow melt and drip processes will likely become more common. However, since melt and drip of canopy snow is not well represented in current hydrological models, additional investigations are required to better understand these processes that are expected to become more prevalent with climate change. 

Climate change is also expected to have profound effects on forest ecology which presents problems for snow interception parameterisations that have coefficients derived from individual sites and applied globally without change. Large portions of the boreal forest have been found to be experiencing significant ecosystem change and has been attributed to browning from drought [@Dai2013], disease [@Ruess2021], insects [@Kurz2008], and fire [@Kasischke2006; @Kasischke2010]. Following these disturbances greening occurs, which increases forest density in the disturbed regions [@Sulla2018; @Ju2016; @Keenan2018; @Lantz2019]. Across the globe, mountain ecosystems have also observed rapid vegetation changes with evidence for an increase in warm-adapted species and reduction in cold-adapted species [@Gottfried2012]. These changes in temperature, precipitation and forest ecology are expected to have strong modifications on how intercepted snow is partitioned through sublimation, unloading and drip [@He2021]. This introduces uncertainty when parameterisations and coefficients, developed in a 20th century climate are applied to new climates and forest ecosystems that are arising in the 21st century.

These rapid changes of climate and forest ecology illustrate the pressing need to assess whether existing snow interception parameterisations are suitable for the uncalibrated application across diverse forest ecosystems and climates. If the physical processes important for snow interception are adequately represented in existing parameterisations, then more accurate model performance should be expected across diverse environments [@Pomeroy2013a; @Clark2016]. Therefore, the theory and assumptions of existing parameterisations need to be tested across variable climate and forest structures to ensure their applicability in these environments and whether modification is required. This will ascertain where current processes representations work and what processes are not presently included that should be. 

To achieve this, first the theoretical underpinnings and assumptions of existing parameterisations must be tested across forests that have variable canopy cover, wind speed, and air temperatures. Previous work by @Gelfan2004, has shown the [@Hedstrom1998; @Pomeroy1998b] snow interception parameterisation to perform well when coupled with the @Storck2002 wet-snow unloading and drip algorithm in northwestern Russia which frequently transitioned from 0 to -20°C in canopy cover ranging from 0.6 to 0.7. However, aside from the @Gelfan2004 study, the evaluation of snow interception parameterisations has been limited and few studies have investigated their performance in mountain forests. Therefore, in-situ measurements are needed of canopy snow interception and the subsequent ablation of canopy snow in differing mountain climates (warm, transitional and cold) and forest structures (sparse to dense). These measurements will be used to test the theoretical underpinnings of existing parameterisations for application across environments that have differing processes important for forest snow redistribution.

To assess the performance of existing parameterisations, a modular and flexible model platform will be used to simulate the individual mass and energy balance fluxes and states of the canopy and snowpack. The advantage of modular and flexible model platforms is that it is possible to adjust process parameterisations while holding the remaining model parameters and mass and energy conservation equations constant [@Pomeroy2007; @Clark2015]. The Cold Regions Hydrological Model platform [CRHM, @Pomeroy2007] will be used in this study as its snow interception algorithms have been successfully used in previous studies [@Ellis2010; @Pomeroy2012; @Rasouli2019b; @Sanmiguel2022] to assess snow interception processes in Spain, USA, and Canada (Alberta, Idaho, Yukon, and Saskatchewan). Since CRHM is a process-based model agnostic platform, individual state and flux variables including subcanopy snow accumulation, snow interception, sublimation, unloading, drip, and wind redistribution can be evaluated using differing parameterisations. To evaluate existing parameterisations of snow interception [@Hedstrom1998;@Storck2002;@Andreadis2009], and ablation of canopy snow by unloading [@Ellis2010; @Roesch2001; @Andreadis2009], drip [@Storck2002;@Andreadis2009; @Ellis2010] and sublimation [@Pomeroy1998b] CRHM will be run at four mountain forest research basins across western Canada and compared to in-situ observations. Insights gained from an evaluation of the theoretical underpinnings and assumptions on new and archival field observations will help inform the improvement of existing snow interception process representations. Modifications to the existing parameterisations to better represent snow interception processes will be validated on the associated change in error in simulated and observed forest snow accumulation and streamflow.

## Research Gap and Objectives

This subsection of Chapter 1 will describe the overall purpose of the thesis and links the individual thesis objectives to research gaps.

Thesis objectives and research questions

Purpose: To better understand the processes that govern snow accumulation in forested environments.

1. Evaluate the suitability of existing snow interception and ablation parameterisations for application in needleleaf forests with differing canopy structure and meteorology.

   1.1 What are the theoretical underpinnings and assumptions behind existing snow interception and ablation parameterisations?  
   
   1.2 Are the theories and assumptions of existing snow interception parameterisations supported by field measurements collected across diverse canopy structures and meteorological conditions?  
   
   1.3 Are the theories and assumptions of existing canopy snow ablation parameterisations supported by field measurements collected across varying meteorological conditions?  

2. Determine how new snow interception and ablation parameterisations could enhance the representation of processes important for subcanopy snow accumulation.

   2.1 How can the use of novel snow interception parameterisations enhance simulations of snow accumulation in forests with differing tree species, canopy structures, and meteorological conditions?  


## Organization of Chapters

This thesis contains 5 chapters, the first chapter includes an introduction and research plan while, the remaining chapters 2-5 each correspond to a journal article which aims to answer one of the research questions.

\pagebreak

# The Theoretical Underpinnings of Existing Snow Interception and Ablation parameterisations

Manuscript Status: The contents of this chapter have been compiled from a advanced review article published in the journal *WIREs Water*.

Citation: **Cebulski, A. C.**, & Pomeroy, J. W. (2025). Theoretical Underpinnings of Snow Interception and Canopy Snow Ablation Parameterisations. WIREs Water, 12, e70010. https://doi.org/10.1002/wat2.70010

Role in thesis: This paper is an advanced review article and corresponds to objective 1, research question 1.1 of the thesis. This advanced review will provide the context necessary for interpreting whether the theories and assumptions of existing parameterisations are true for the field observations collected in this study in the second part of objective 1.

Author Contribution: A. C. Cebulski: conceptualization (lead), investigation (equal), software (lead), visualization (lead), writing – original draft (lead), writing review and editing (equal). J. W. Pomeroy: conceptualization (supporting), funding acquisition (lead), investigation (equal), project administration (lead), resources (lead), supervision (lead), writing review and editing (equal).

```{r child = 'chapters/02-review-paper.qmd'}
```

\pagebreak

# Snow Interception Relationships with Meteorology and Canopy Density

Manuscript status: The contents of this chapter have been compiled from a research article published in the journal *Hydrological Processes*. 

Citation: **Cebulski, A. C.**, & Pomeroy, J. W. (2025). Snow Interception Relationships With Meteorology and Canopy Density. Hydrological Processes, 39(4), e70135. https://doi.org/10.1002/hyp.70135

Role in thesis: This journal article aims to answer research question 1.2 of the thesis. This question will be answered through analysis of observations of interception from a study site few researchers have focused on, a subalpine discontinuous forest and contrast these observations with existing theory developed in maritime and continental climates.

Author Contribution: Alex Cebulski conducted in-situ data collection, conceptualized research plan with comments from supervisor and thesis committee members, and lead author in collaboration with supervisor. 

```{r child = 'chapters/03-snow-int-paper.qmd'}
```

\pagebreak

# Processes Governing the Ablation of Intercepted Snow

Manuscript status: This chapter has been submitted as a research article to the journal *Water Resources Research* on August 19, 2025 and is currently under review.

Citation: **Cebulski, A. C.**, & Pomeroy, J. W. (under review). Processes Governing the Ablation of Intercepted Snow. Water Resources Research.

Role in thesis: This journal article answers research question 1.3 of the thesis. It will present analysis of canopy snow ablation observations from Fortress Mountain Research basin collected over the 2022 and 2023 water years and contrast these against existing theories.

```{r child = 'chapters/04-ablation-paper.qmd'}
```

\pagebreak

# Evaluation of a New Needleleaf Forest Snowpack Model for Diagnosing Snow Accumulation Regimes in Western and Northern Canada

Manuscript status: Analysis is complete and manuscript is currently in progress. Anticipated submission is to the journal *Hydrological Processes* Fall, 2025.

Role in thesis: This journal article corresponds to research question 2.1 of the thesis, the model implementation and testing phase. New parameterisations and process understanding of snow interception and ablation processes from chapters 1, 2 and 3 will incorporated into the Cold Regions Hydrological Model (CRHM) platform to model forest snow accumulation at four research basins in western Canada. The updated parameterisations will be evaluated by including them in an updated CRHM canopy module. Simulated SWE using this updated module will be compared to observed SWE at a point within the forest of each research basin. 

```{r child = 'chapters/05-model-eval-paper.qmd'}
```

\pagebreak

# Conclusions

```{r child = 'chapters/06-synth-conclusion.qmd'}
```

\pagebreak

# References

::: {#refs}
:::

<!-- # Appendix for Chapter 2 {#sec-appendix-c1} -->

<!-- ```{r child = 'chapters/02-review-paper-appdx.qmd'} -->
<!-- ``` -->

<!-- # Appendix for Chapter 3 -->

<!-- ```{r child = 'chapters/03-snow-int-paper-appdx.qmd'} -->
<!-- ``` -->
