---
title: "Advancing Snow Accumulation Models in Needleleaf Forests"
author: "Alex Cebulski"
date: today
date-format: long
format: 
  # html
  # pdf
  docx:
    reference-doc: ../config/thesis-word-template.docx
    number-sections: true
    toc: false
editor: source
bibliography: ../../papers/bibtex/library.bib
csl: '../../papers/bibtex/csl-styles/apa-disable-disambiguate-add-names.csl'
---

# Introduction

## Background

This subsection of Chapter 1 will introduce background information, motivation, hydrological significance of the study topic, research gaps, and methods used in the study.

## Research Gap and Objectives

This subsection of Chapter 1 will describe the overall purpose of the thesis and links the individual thesis objectives to research gaps.

Thesis objectives and research questions

Purpose: To better understand the processes that govern snow accumulation in forested environments.

1. Evaluate the suitability of existing snow interception and ablation parameterizations for application in needleleaf forests with differing canopy structure and meteorology.

    1.1 What are the theoretical underpinnings and assumptions behind existing snow interception and ablation parameterizations?
    
    1.2 Are the theories and assumptions of existing snow interception parameterizations supported by field measurements collected across diverse canopy structures and meteorological conditions?
    
    1.3 Are the theories and assumptions of existing canopy snow ablation parameterizations supported by field measurements collected across varying meteorological conditions?

2. Determine how new snow interception and ablation parameterizations could enhance the representation of processes important for subcanopy snow accumulation.

    2.1 How can the use of novel snow interception parameterizations enhance simulations of snow accumulation in forests with differing tree species, canopy structures, and meteorological conditions?

## Organization of Chapters

This thesis contains 5 chapters, the first chapter includes an introduction and research plan while, the remaining chapters 2-5 each correspond to a journal article which aims to answer each of the research questions.

# The Theoretical Underpinnings of Existing Snow Interception and Ablation Parameterizations

Manuscript Status: Invited for submission to the journal WIREs Water and is currently under review.

Role in thesis: This paper is an advanced review article and corresponds to objective 1, research question 1.1 of the thesis. This advanced review will provide the context necessary for interpreting whether the theories and assumptions of existing parameterizations are true for the field observations collected in this study in the second part of objective 1.

Author Contribution: Conducted literature review, committee members provided edits... 

```{r child = 'chapters/02-review-paper.qmd'}
```

# Snow Interception Relationships with Meteorology and Canopy Structure in a Subalpine Forest

Manuscript status: In preparation for submission to the special issue "Snow to Flow" in the Journal Hydrological Processes. 

Role in thesis: This journal article aims to answer research question 1.2 of the thesis. This question will be answered through analysis of observations of interception from a study site few researchers have focused on, a subalpine discontinuous forest and contrast these observations with existing theory developed in maritime and continental climates.

Author Contribution: Conducted in-situ data collection, conceptualized research plan with comments from supervisor and thesis committee members, and lead author in collaboration with supervisor. 

```{r child = 'chapters/03-snow-int-paper.qmd'}
```

# The Influence of Meteorology on Canopy Snow Ablation Processes

Manuscript status: This article is preparation for submission to the journal *Water Resources Research* by October 31, 2024. The results from this paper were presented at INARCH and AGU 2023.

Role in thesis: This journal article answers research question 1.3 of the thesis. It will present analysis of canopy snow ablation observations from Fortress Mountain Research basin collected over the 2022 and 2023 water years and contrast these against existing theories.

```{r child = 'chapters/04-ablation-paper.qmd'}
```

# An Evaluation of New Snow Interception and Ablation Parameterizations in Three Mountain Forests in western Canada

Manuscript status: Model setup and simulations in progress writing has not started. Anticipated submission is to the journal *Water Resources Research* March, 2025.

Role in thesis: This journal article corresponds to research question 2.1 of the thesis, the model implementation and testing phase. New parameterizations and process understanding of snow interception and ablation processes from chapters 1, 2 and 3 will incorporated into the Cold Regions Hydrological Model (CRHM) platform to model forest snow accumulation at four research basins in western Canada. The updated parameterizations will be evaluated by including them in an updated CRHM canopy module. Simulated SWE using this updated moduled will be compared to observed SWE within the forested portion of each basin. 

\pagebreak

# References

::: {#refs}
:::

# Appendix for Chapter 2 {#sec-appendix-c1}

```{r child = 'chapters/02-review-paper-appdx.qmd'}
```

# Appendix for Chapter 3

```{r child = 'chapters/03-snow-int-paper-appdx.qmd'}
```
